---
layout: page
title: Programming with R
subtitle: Introduction to RStudio
minutes: 15
---

<!-- ```{r, include=FALSE, echo=FALSE} -->
<!-- source("tools/chunk-options.R") -->
<!-- ``` -->

> ## Objectives {.objectives}
> * Get familiar with RStudio interface
> * Understand the difference between script file and console
> * Demonstrate useful shortcuts
> * Notions of variables, workspace, working directory and function


Let's start by learning about our tool.

* Point to the different panels: Console, Scripts, Environments, Plots
* Code and workflow are more reproducible if we can document everything that we do.
* Our end goal is not just to "do stuff" but to do it in a way that anyone can
  easily and exactly replicate our workflow and results.
* The best way to achieve this is to write scripts. 

### Interacting with R

There are two main ways of interacting with R: using the console or by using
script files (plain text files that contain your code).

The console window (in RStudio, the bottom left panel) is the place where R is
waiting for you to tell it what to do, and where it will show the results of a
command.  You can type commands directly into the console, but they will be
forgotten when you close the session. It is better to enter the commands in the
script editor, and save the script. This way, you have a complete record of what
you did, you can easily show others how you did it and you can do it again later
on if needed. You can copy-paste into the R console, but the Rstudio script
editor allows you to 'send' the current line or the currently selected text to
the R console using the `Ctrl-Enter` shortcut.

At some point in your analysis you may want to check the content of variable or
the structure of an object, without necessarily keep a record of it in your
script. You can type these commands directly in the console. RStudio provides
the `Ctrl-1` and `Ctrl-2` shortcuts allow you to jump between the script and the
console windows.

If R is ready to accept commands, the R console shows a `>` prompt. If it
receives a command (by typing, copy-pasting or sent from the script editor using
`Ctrl-Enter`), R will try to execute it, and when ready, show the results and
come back with a new `>`-prompt to wait for new commands.

If R is still waiting for you to enter more data because it isn't complete yet,
the console will show a `+` prompt. It means that you haven't finished entering
a complete command. This is because you have not 'closed' a parenthesis or
quotation. If you're in RStudio and this happens, click inside the console
window and press `Esc`; this should help you out of trouble.


###  The working directory

A directory (folder) that lives on your hard drive: `getwd` and
`setwd`. To list the files in the workding directory, use `dir()`.

### Create, list and delete variables:

**Variables** are named containers for data. This data can be simple
(a single number ) or quite big and complex. `<-` is the **assignment
operator**. It assigns values on the right to objects on the left. So,
after executing `x <- 3`, the value of `x` is `3`. The arrow can be
read as 3 **goes into** `x`.  You can also use `=` for assignments but
not in all contexts so it is good practice to use `<-` for
assignments. `=` should only be used to specify the values of
arguments in functions, see below.

```{r, echo=TRUE}
x <- 1
y = 2
x
y
```

The variables that were created live in the **global environment** or
**workspace**, which is a space for data in R (i.e. not saved on your
hard drive; if you close R, all variables will be lost). To list the
content of the workspace, use `ls()`. To remove a variable, use
`rm(varname)`, where `varname` is the name of the variable to be
removed.

```{r, echo=TRUE}
ls()
rm(x)
ls()
```

> ## Tip {.callout}
>
> Be consistent in using either `<-` or `=` as assignment operator.

In RStudio, typing `Alt + -` (push `Alt`, the key next to your space bar at the
same time as the `-` key) will write ` <- ` in a single keystroke.

### Functions

**Functions** are mini-programs that perform a set of defined
  operations on specific inputs (the function **arguments** or
  **parameters**). The result of these operations is the output of the
  function. Functions have unique names and can be easily recognised
  by the parenthesis, like `ls()` above. The inputs are defined inside
  of the parenthesis.

```{r}
## only variables that have a 'y' in their name
ls(pattern = "y")
```

### Finding help

Any function or data variables that are distributed by R or any of its
valid packages is fully documented. The documentation can be accessed
in two equivalent ways:

```{r, eval=FALSE}
?setwd
help("setwd")
```

Some packages also provide high-level documentation that provide an
overview of the package. These documents are called vignettes:

```{r, eval=FALSE}
vignette(package = "Matrix")
vignette("Introduction", package = "Matrix")
```

### Quitting

To quit R, use the

```{r, eval=FALSE}
q()
```

function, which will ask if you want to store all the variables in
your workspace in a file called (by default) `.RData` in the working
directory. Inidivual variables can be saved with `save` and later on
load back again with `load`.

```{r, eval=TRUE}
save(y, file = "y.RData")
rm(y)
load("y.RData")
y
```

> ## Tip {.callout}
>
> It is better to store specific variables of interest rather than
> saving the full complete workspace.

```{r, echo=FALSE}
unlink("y.RData")
```

### Commenting

Use `#` signs to comment. Comment liberally in your R scripts. Anything to the
right of a `#` is ignored by R.
