---
layout: page
title: Programming with R
subtitle: Making choices
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/04-cond-")
```

> ## Objectives {.objectives}
>
> *   Write conditional statements with `if` and `else`.
> *   Correctly evaluate expressions containing `&` ("and") and `|` ("or").

Our previous lessons have shown us how to manipulate data, define our own functions, and repeat things.
However, the programs we have written so far always do the same things, regardless of what data they're given.
We want programs to make choices based on the values they are manipulating.
 
## Conditionals

In order to update our function to decide between saving or not, we need to write code that automatically decides between multiple options.
The tool R gives us for doing this is called a [conditional statement](../../gloss.html#conditional-statement), and looks like this:

```{r, results='hold'}
num <- 37
if (num > 100) {
  print("greater")
} else {
  print("not greater")
}
print("done")
```

The second line of this code uses an `if` statement to tell R that we want to make a choice.
If the following test is true, the body of the `if` (i.e., the lines in the curly braces underneath it) are executed.
If the test is false, the body of the `else` is executed instead.
Only one or the other is ever executed:

<img src="fig/python-flowchart-conditional.svg" alt="Executing a Conditional" />

In the example above, the test `num > 100` returns the value `FALSE`, which is why the code inside the `if` block was skipped and the code inside the `else` statment was run instead.

```{r}
num > 100
```

And as you likely guessed, the opposite of `FALSE` is `TRUE`.

```{r}
num < 100
```

Conditional statements don't have to include an `else`.
If there isn't one, R simply does nothing if the test is false:

```{r}
num <- 53
if (num > 100) {
  print("num is greater than 100")
}
```

We can also chain several tests together when there are more than two options.
This makes it simple to write a function that returns the sign of a number:

```{r}
sign <- function(num) {
  if (num > 0) {
    return(1)
  } else if (num == 0) {
    return(0)
  } else {
    return(-1)
  }
}

sign(-3)
sign(0)
sign(2/3)
```

Note that the test for equality uses two equal signs, `==`.

> ## Tip {.callout}
>
> Other tests include greater than or equal to (`>=`), less than or equal to
> (`<=`), and not equal to (`!=`).

We can also combine tests.
An ampersand, `&`, symbolizes "and".
A vertical bar, `|`, symbolizes "or".
`&` is only true if both parts are true:

```{r}
if (1 > 0 & -1 > 0) {
    print("both parts are true")
} else {
  print("at least one part is not true")
}
```

while `|` is true if either part is true:

```{r}
if (1 > 0 | -1 > 0) {
    print("at least one part is true")
} else {
  print("neither part is true")
}
```

In this case, "either" means "either or both", not "either one or the other but not both".

> ## Challenge - Using conditions to change behaviour {.challenge}
>
> Write a function that converts between C, F and K temperatures.  The
> function will need three arguments: (1) temperature (`t`), a
> numeric, (2) from unit (`from`), a character and (3) to unit
> (`to`). They will need default values, e.g `t = 0`, `from = "C"` and
> `to = "K"`.
>
> Conversion formulas: `F = 9/5 C + 32` and `K = C + 273.15`


<div class="keypoints" markdown="1">
#### Key Points

* Use `if (condition)` to start a conditional statement, `else if
  (condition)` to provide additional tests, and `else` to provide a
  default.
* The bodies of conditional statements must be surrounded by curly
  braces `{ }`.
* Use `==` to test for equality.
* `X & Y` is only true if both X and Y are true.
* `X | Y` is true if either X or Y, or both, are true.  </div>

#### Next Steps

We have now seen the basics of interactively building R code. We will
not move on on more plotting and conclude with reproducible report
generation.

