---
layout: page
title: Programming with R
subtitle: Making choices
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/04-cond-")
```

> ## Objectives {.objectives}
>
> *   Save plot(s) in a pdf file.
> *   Write conditional statements with `if` and `else`.
> *   Correctly evaluate expressions containing `&` ("and") and `|` ("or").

Our previous lessons have shown us how to manipulate data, define our own functions, and repeat things.
However, the programs we have written so far always do the same things, regardless of what data they're given.
We want programs to make choices based on the values they are manipulating.

### Saving plots to a file

So far, we have built a function `analyze` to plot summary statistics of the inflammation data:

```{r analyze}
analyze <- function(filename) {
  # Plots the average, min, and max inflammation over time.
  # Input is character string of a csv file.
  dat <- read.csv(file = filename, header = FALSE)
  avg_day_inflammation <- apply(dat, 2, mean)
  plot(avg_day_inflammation)
  max_day_inflammation <- apply(dat, 2, max)
  plot(max_day_inflammation)
  min_day_inflammation <- apply(dat, 2, min)
  plot(min_day_inflammation)
}
```

And also built the function `analyze_all` to automate the processing of each data file:

```{r analyze_all}
analyze_all <- function(pattern) {
  # Runs the function analyze for each file in the current working directory
  # that contains the given pattern.
  filenames <- list.files(pattern = pattern)
  for (f in filenames) {
    analyze(f)
  }
}
```

While these are useful in an interactive R session, what if we want to send our results to our collaborators?
Since we currently have 12 data sets, running `analyze_all` creates 36 plots.
Saving each of these individually would be tedious and error-prone.
And in the likely situation that we want to change how the data is processed or the look of the plots, we would have to once again save all 36 before sharing the updated results with our collaborators.

Here's how we can save all three plots of the first inflamation data set in a pdf file:

```{r, results='hide'}
pdf("inflammation-01.pdf")
analyze("data/inflammation-01.csv")
dev.off()
```

The function `pdf` redirects all the plots generated by R into a pdf file, which in this case we have named "inflammation-01.pdf".
After we are done generating the plots to be saved in the pdf file, we stop R from redirecting plots with the function `dev.off`.

> ## Tip {.callout}
>
> If you run `pdf` multiple times without running `dev.off`, you will save plots to the most recently opened file.
> However, you won't won't be able to open the previous pdf files because the connections were not closed.
> In order to get out of this situation, you'll need to run `dev.off` until all the pdf connections are closed.
> You can check your current status using the function `dev.cur`.
> If it says "pdf", all your plots are being saved in the last pdf specified.
> If it says "null device" or "RStudioGD", the plots will be visualized normally.

We can update the `analyze` function so that it always saves the plots in a pdf.
But that would make it more difficult to interactively test out new changes.
It would be ideal if `analyze` would either save or not save the plots based on its input.

### Conditionals

In order to update our function to decide between saving or not, we need to write code that automatically decides between multiple options.
The tool R gives us for doing this is called a [conditional statement](../../gloss.html#conditional-statement), and looks like this:

```{r, results='hold'}
num <- 37
if (num > 100) {
  print("greater")
} else {
  print("not greater")
}
print("done")
```

The second line of this code uses an `if` statement to tell R that we want to make a choice.
If the following test is true, the body of the `if` (i.e., the lines in the curly braces underneath it) are executed.
If the test is false, the body of the `else` is executed instead.
Only one or the other is ever executed:

<img src="fig/python-flowchart-conditional.svg" alt="Executing a Conditional" />

In the example above, the test `num > 100` returns the value `FALSE`, which is why the code inside the `if` block was skipped and the code inside the `else` statment was run instead.

```{r}
num > 100
```

And as you likely guessed, the opposite of `FALSE` is `TRUE`.

```{r}
num < 100
```

Conditional statements don't have to include an `else`.
If there isn't one, R simply does nothing if the test is false:

```{r}
num <- 53
if (num > 100) {
  print("num is greater than 100")
}
```

We can also chain several tests together when there are more than two options.
This makes it simple to write a function that returns the sign of a number:

```{r}
sign <- function(num) {
  if (num > 0) {
    return(1)
  } else if (num == 0) {
    return(0)
  } else {
    return(-1)
  }
}

sign(-3)
sign(0)
sign(2/3)
```

Note that the test for equality uses two equal signs, `==`.

> ## Tip {.callout}
>
> Other tests include greater than or equal to (`>=`), less than or equal to
> (`<=`), and not equal to (`!=`).

We can also combine tests.
An ampersand, `&`, symbolizes "and".
A vertical bar, `|`, symbolizes "or".
`&` is only true if both parts are true:

```{r}
if (1 > 0 & -1 > 0) {
    print("both parts are true")
} else {
  print("at least one part is not true")
}
```

while `|` is true if either part is true:

```{r}
if (1 > 0 | -1 > 0) {
    print("at least one part is true")
} else {
  print("neither part is true")
}
```

In this case, "either" means "either or both", not "either one or the other but not both".

> ## Challenge - Using conditions to change behaviour {.challenge}
>
> Write a function that converts between C, F and K temperatures.  The
> function will need three arguments: (1) temperature (`t`), a
> numeric, (2) from unit (`from`), a character and (3) to unit
> (`to`). They will need default values, e.g `t = 0`, `from = "C"` and
> `to = "K"`.
>
> Conversion formulas: `F = 9/5 C + 32` and `K = C + 273.15`


<div class="keypoints" markdown="1">
#### Key Points

* Use `if (condition)` to start a conditional statement, `else if
  (condition)` to provide additional tests, and `else` to provide a
  default.
* The bodies of conditional statements must be surrounded by curly
  braces `{ }`.
* Use `==` to test for equality.
* `X & Y` is only true if both X and Y are true.
* `X | Y` is true if either X or Y, or both, are true.  </div>

#### Next Steps

We have now seen the basics of interactively building R code. We will
not move on on more plotting and conclude with reproducible report
generation.

